Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-07-16 Thesis/Paper_II.tex
Semi-final presubmission draft

\chapter{Cosmological Bar Formation: Nature vs Nurture}

%\last_written_page

% Abstract of the paper
\section{Abstract}

We investigate the connection between the vertical structure of
stellar discs and the formation of bars using high-resolution
simulations of galaxies in isolation and in the cosmological context.
In particular, we simulate a suite of isolated galaxy models that have
the same Toomre $Q$ parameter and swing amplification parameter but
that differ in the vertical scale height and velocity dispersion.  We
find that the onset of bar formation occurs more slowly in models with
thicker discs.  Moreover, thicker discs and also discs evolved in
simulations with larger force softening also appear to be more
resilient to buckling, which acts to regulate the length and strength
of bars.  We also simulate disc-halo systems in the cosmological
environment using a disc-insertion technique developed in a previous
paper.  In this case, bar formation is driven by the stochastic
effects of a triaxial halo and subhalo-disc interactions and the
initial growth of bars appears to be relatively insensitive to the
thickness of the disc.  On the other hand, thin discs in cosmological
halos do appear to be more susceptible to buckling than thick ones and
therefore bar strength correlates with disc thickness as in the
isolated case.  More to the point, one can form discs in cosmological
simulations with relatively weak bars or no bars at all provided the
discs as thin as the discs we observe and the softening length is
smaller than the disc scale height.

@@

The problem of bar formation in disc galaxies tests our understanding
of cosmological structure formation and galactic dynamics.  In
principle, theories of galaxy formation should yield predictions for
the fractional distribution of bars in terms of their strength,
length, and pattern speed.  While it is often difficult to make
precise, quantitative statements about bars from observations, general
properties of their distribution have emerged (See
\citet{sellwood1993}, \citet{Sellwood2013} and \citet{BT} for
reviews).  Roughly 30-40 per cent of disc galaxies exhibit strong
bars, that is bars that dominate the disc luminosity.  Another 20 per
cent or more have relatively weak bars.  The bar fraction appears to
increase with time.  Approximately one tenth of disc galaxies between
$0.5 \le z \le 2$ have visually identifiable strong bars, which is a
factor of 3-4 smaller than the fraction in the local Universe (see
\citet{simmons2014} and references therein).  The bar fraction also
varies with galaxy type.  \citet{masters2010} find that $70\pm 5$ per
cent of the so-called passive red spirals have bars as compared to a
$25\pm 5$ per cent bar fraction for blue spirals.  Since the red
spirals are interpreted as old galaxies that have used up their
star-forming gas, this result is consistent with a bar fraction that
increases with time.  The upshot of these observations is that in
terms of bars, disc galaxies in the local Universe divide into three
roughly equal bins: those with strong bars, those with weak bars, and
those with no detectable bar.  These observations suggest that bars are 
capable of forming at a wide range of cosmic times, but once formed,
are difficult to destroy.

Intuition tells us that properties of a bar should depend on the
properties of its host galaxy and the environment in which that
galaxy lives.  Theoretical arguments indicate that cold, thin discs
are susceptible to local ``Toomre'' instabilities.  Furthermore, discs
that are strongly self-gravitating, that is discs that contribute the
bulk of the gravitational force required to maintain their rotational
motion, are susceptible to global instabilities.  Thus, one can
construct initially axisymmetric galaxy models that form bars with
vastly different properties (or no bars at all) by changing the
internal disc dynamics or trading off disc mass for bulge and halo
mass.  The implication is that the distribution of bars provides an
indirect means for inferring a disc's kinematics and mass-to-light
ratio as well as the distribution of mass in a galaxy's dynamically
``hot'' components, namely its bulge and dark matter halo.

A galaxy's ability to resist local instabilities is typically
expressed in terms of the (kinetic) Toomre $Q$-parameter
\citep{ToomreParameter} while its ability to resist global
instabilities is encapsulated in the swing-amplification $X$-parameter
\citep{GoldreichTremaine1978,GoldreichTremaine1979}.  Both parameters
are defined so that large values imply a more stable disc.  The
hypothesis that they determine a galaxy's susceptibility to bar
formation has been tested by simulations of isolated, idealized galaxy
models \citep{PeeblesOstriker1973, ZangHohlBars1978,
  CombesSandersBars1981,Sellwood1981}.  Typically, the initial galaxy
is represented by an N-body (Monte Carlo) realization of an equilibrium solution to
the collisionless Boltzmann equation comprising a disc, dark matter
halo, and often, a bulge.  Equilibrium does not imply
stability, and a galaxy can develop spiral structure and a bar through
instabilities that are seeded by shot noise
\citep{EfstathiouShotNoise} and amplified by feedback loops such as
swing amplification \citep{Sellwood2013}. A common way to suppress the mechanisms
which give rise to these effects is by increasing either $Q$ or $X$.  For
example, in dynamically warm discs (that is, discs with high $Q$)
perturbations are randomized on timescales short enough to prevent the
feedback loop from starting \citep{AthanassoulaSellwood1986}.
Likewise, submaximal discs, that is, discs with high $X$, avoid the
bar instability presumably because the disc lacks the self-gravity to
drive the bar instability \citep{EfstathiouShotNoise,
  ChristodoulouStability1995, Sellwood2013}.

@@

Of course, galaxies are neither isolated nor born as axisymmetric,
equilibrium systems. In these idealized systems, instabilities can only come from shot noise, whereas this is is not true in a complicated dynamical environment. As such, bar formation may be very different in idealized
galaxies as compared with galaxies in a cosmological setting.  For
example, halo substructure in the form of satellite galaxies and dark
matter subhaloes can pass through and perturb the disc.
\citet{gauthier2006}, \citet{kazantzidis2008}, and
\citet{dbs2009} showed that an apparently stable disc galaxy
model can develop a bar when a fraction of the ``smooth'' halo is
replaced by substructure in the form of subhaloes.  The effect is
stochastic; subhalo-triggered bar formation seems to require subhaloes
whose orbits take them into the central regions of the disc in a
prograde sense.  More recently, \citet{purcell2011} showed that
Sagittarius dwarf alone could have been responsible for the Milky
Way's spiral structure and bar.

@@

Though the \citet{YurinSpringelStellarDisks} models vary in $Q$ and
$X$ they share two important properties.  First, the ratio of the
radial and vertical velocity dispersion is set to unity throughout the
disc.  Second, the ratio of the vertical and radial scale lengths is
fixed at $0.2$, which is roughly a factor of two larger than that of
the Milky Way's thin disc.  In addition, the softening length in their
simulations is fixed at $680\,{\rm pc}$.  Thus, the discs in their
simulations are thicker and (vertically) warmer than what one might
expect for Milky Way-like galaxies.  The results from
\citet{Klypin2009} suggest that these properties rather than (or
together with) halo substructure and triaxiality are responsible for
the preponderance of strong bars in the
\citet{YurinSpringelStellarDisks} models.

@@

We begin in section \S \ref{sec:ics} by discussing the dimensionless
parameters that arise when one constructs equilibrium disc galaxy
models.  These include the aforementioned $Q$ and $X$ parameters as
well as the ratio of the vertical and radial velocity dispersion and
the ratio of the vertical and radial scale lengths.  We also discuss
possible effects of gravitational softening.  In
\ref{sec:thick_discs_suppress} we describe our sequence of simulations
and discuss how they fit in with previous work.  We discuss results
for our isolated galaxy simulations in \S \ref{sec:isolated} and our
cosmological ones in \S \ref{sec:cosmo}.  We conclude with a
discussion of the implications of our results in \S
\ref{sec:conclusions}.

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/rotation_curves_two_panel.eps}
	\caption{Rotation curve decomposition for our models. Total
          rotation curves are shown as solid lines while the separate
          contributions from the disc and halo are shown as dashed and
          dot-dashed curves, respectively.  Blue curves in the top
          panel are for the isolated galaxy simulations with
          \textsc{GalactICS} initial conditions while the green curves
          are for the simulations C.I.Ag run with \textsc{AGAMA}
          initial conditions.  Bottom panel shows initial rotation
          curve decomposition for the runs D.I and
          E.II.}\label{fig:rcs}
\end{figure}

\begin{table*}
\begin{tabular}{l l l l l l l l l l l l}
\hline
 & $M_d$ & $R_d$ & $V_c$ & $\sigma_R$ & $z_d/R_d$  & $\sigma_R/\sigma_z$  & $X$ &  $Q$ & $\rho_h/\rho_0$ & $\epsilon$ \\ 
\hline
A.I     & 3.49 & 2.50 & 216 & 25.3 & 0.10 & 1.27 & 2.34 & 1.00 & 0.14 & 0.15\\
A.II    & 3.49 & 2.50 & 216 & 25.3 & 0.10 & 1.27 & 2.34 & 1.00 & 0.14 & 0.50\\
B.I     & 3.49 & 2.50 & 213 & 25.3 & 0.20 & 0.97 & 2.34 & 1.00 & 0.28 & 0.15\\
B.II    & 3.49 & 2.50 & 213 & 25.3 & 0.20 & 0.97 & 2.34 & 1.00 & 0.28 & 0.50\\
C.I     & 3.49 & 2.50 & 208 & 25.3 & 0.40 & 0.77 & 2.34 & 1.00 & 0.50 & 0.15\\
C.I.Ag  & 3.49 & 2.50 & 216 & 25.3 & 0.40 & 0.77 & 2.34 & 1.00 & 0.50 & 0.15\\
D.IV    & 5.82 & 3.70 & 245 & 25.2 & 0.10 & 1.14 & 2.45 & 1.00 & 0.29 & 0.18\\
E.IV    & 5.82 & 3.70 & 245 & 27.4 & 0.25 & 0.72 & 2.45 & 1.00 & 0.73 & 0.74\\
\hline
YS15.A5 & 5.00 & 3.00 & 263 & 30.7  & 0.2 & 1.00 & 3.22 & 1.38 & 0.21 & 0.68\\
YS15.B5 & 5.00 & 3.00 & 211 & 26.6  & 0.2 & 1.00 & 2.06 & 0.96 & 0.11 & 0.68\\
YS15.C5 & 5.00 & 3.00 & 270 & 30.3  & 0.2 & 1.00 & 3.31 & 1.42 & 0.23 & 0.68\\
YS15.D5 & 5.00 & 3.00 & 236 & 26.6  & 0.2 & 1.00 & 2.58 & 1.12 & 0.16 & 0.68\\
YS15.E5 & 5.00 & 3.00 & 233 & 27.1  & 0.2 & 1.00 & 2.58 & 1.11 & 0.15 & 0.68\\
YS15.F5 & 5.00 & 3.00 & 219 & 27.0  & 0.2 & 1.00 & 2.22 & 1.02 & 0.11 & 0.68\\
YS15.G5 & 5.00 & 3.00 & 227 & 28.2  & 0.2 & 1.00 & 2.45 & 1.09 & 0.13 & 0.68\\
YS15.H5 & 5.00 & 3.00 & 244 & 28.6  & 0.2 & 1.00 & 2.85 & 1.21 & 0.16 & 0.68\\
\hline
G06     & 7.77 & 5.57 & 226 & 17.1  & 0.06  & 1.80 & 2.78 & 1.43 & 0.10 & 0.15 \\
\hline
%$M_d$ & $7.2 \times 10^{10} M_\odot$\\
%$N_d$ & $10^6$\\
%$R_{d,0}$ & 3.7 kpc\\
%$N_{r}$ & $4096^3$
\end{tabular}
\caption{Summary of parameters for the simulations considered in this
  paper, the disk-halo simulations considered in
  \citet{YurinSpringelStellarDisks} (labeled YS15) and the
  \citet{gauthier2006} (G06).  $M_d$ is the final disc mass in units of
  $10^{10}\,M_\odot$ , $R_d$ is the disc scale radius in units of
  ${\rm kpc}$, and $V_c$ and $\sigma_R$ are the circular speed and
  radial velocity dispersion in units of ${\rm km\,s^{-1}}$ and
  evaluated at $R_p = 2.2R_d$.  For the disc aspect ratio, we quote
  $z_d/R_d$ where $z_d$ is the ${\rm sech}^2$-scale length.  The
  velocity dispersion ratio $\sigma_R/\sigma_z$, the $X$ and $Q$
  parameters, the ratio of the halo density in the midplane to that of
  the disc, and the logarithmic derivative of the circular speed are
  also measured at $R_p$.  Finally, the softening length $\epsilon$ is
  given in units of ${\rm kpc}$.  Simulations A.III and B.III are the
  same as A.I and B.I except that they are run with vertical motions
  isotropized so as to shut off the buckling
  instability.} \label{tab:sims}
\end{table*}

@@

\begin{figure}
	\centering
        \includegraphics[width=0.9\textwidth]{../figures/falpha.eps}
        \caption{The dimensionless ratio $\sigma_z^2/\pi G\Sigma z_d$ as a
          function of $\rho_h/\rho_0$ for the models considered in
          this paper (stars), the disc-halo models from
          \citet{YurinSpringelStellarDisks} (filled squares) and the
          model from \citet{gauthier2006} (filled triangle).  The
          straight line is the function $f = 1 + \left (2\pi/3\right
          )^{1/2}\rho_h/\rho_0$ discussed in the text.}
\label{fig:falpha}\end{figure}

\begin{figure*}
	\centering
        \includegraphics[width=1.\textwidth]{../figures/qxr.eps}
        \caption{Distribution of simulations considered in this paper
          in the $Q-X$ and the $z_d/R_d-\sigma_R/\sigma_z$ planes.
          Stars are simulations run for this paper (A-E); filled
          squares denote the series of simulations described in
          \citet{YurinSpringelStellarDisks}; the filled triangle denotes
          the simulation of M31 run in \citet{gauthier2006}; open
          pentagons denote the simulations described in
          \citet{WPDGalactICSReference}.
        }\label{fig:QXR}\end{figure*}

\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{../figures/wpd.eps}
	\caption{Strength and length of bars for the simulations
          considered in \citet{gauthier2006}.  The twenty-five models
          span the $Q$-$X$ plane.  Top panel shows the $A_2$ parameter
          while the bottom panel shows the bar length.  Both are
          measured at $5\,{\rm Gyr}$ (the final snapshot of the
          simulations).}
	\label{fig:qxa2}
\end{figure}

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{../figures/isolated_xy_with_agama_with_circles.eps}
	\caption{Surface density maps for isolated galaxy simulations
          at select times. Time proceeds from 0 to 10 Gyr,
          left-to-right, and the models span top-to-bottom in order of
          their appearance in Table \ref{tab:sims}. The overlaid red
          circles have radii $R_p=5.5\,{\rm kpc}$ and 20 kpc.}
	\label{fig:face_on_isolated}
\end{figure*}

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/isolated_a2_vs_t_2rd_weighted.eps}
	\caption{Mean bar strength parameter inside a cylindrical
          radius $R_p$, $A_2(<R_p)$, as a function of time. Curves are
          smoothed in time with a top-hat moving window of width
          $1\,{\rm Gyr}$.  Line colors are blue, red, and green for
          models A, B, and C, respectively.  Results for the
          fiducial runs A.I, B.I, and C.I are shown as solid curves
          while the results for the runs with high softening length,
          A.II and B.II, are shown as dashed curves.  The
          \textsc{AGAMA} model C.I.Ag is shown as a dot-dashed
          curve.} \label{fig:isolated_a2_vs_t}
\end{figure}

\begin{figure*}
	\centering
	\includegraphics[width=1.1\textwidth]{../figures/isolated_r_t_a2_with_agama.eps}
	\caption{Bar strength parameter $A_2$ as a function of radius
          and time.  The trajectory of corotation is shown by the
          dashed red line.} \label{fig:isolated_r_t_a2}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{../figures/isolated_four_panel_zrms.eps}
	\caption{Root mean square height $\zrms$ as a function of
          cylindrical radius $R$ for ten snapshots equally spaced over
          the first $500\,{\rm Myr}$.  Panels are for simulations A.I
          (upper left), A.II (upper right), C.I (lower left) and
          C.I.Ag (lower right).} \label{fig:zrms}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{../figures/isolated_dispersion_evolution_with_agama_zrms.eps}
	\caption{Diagonal components of the velocity dispersion tensor
          and $\zrms$ as a function of $R$ for different snapshots
          between $0$ and $10\,{\rm Gyr}$.  Shown, from to to bottom,
          are profiles for $\zrms$, $\sigma_z$, $\sigma_R$, and $\sigma_\phi$
          for the same size models included in
          Fig.\,\ref{fig:face_on_isolated}.} \label{fig:isolated_dispersions}
\end{figure*}

@@

\subsection{Initial Conditions for Isolated Galaxy Simulations}

>>>>

\chapter{Cosmological Bar Formation: Nature vs Nurture} \label{ch:paper_ii}

\textbf{This chapter contains a reproduction of a paper published in  Monthly Notices of the
Royal Astronomical Society as: Jacob S. Bauer and Lawrence M. Widrow. Can stellar discs in a cosmological setting avoid forming strong bars?. \textit{Monthly Notices of the Royal Astronomical Society}, 476:523-537,
2019. }

\newpage% Abstract of the paper
\section{Abstract}

We investigate the connection between the vertical structure of
stellar discs and the formation of bars using high-resolution
simulations of galaxies in isolation and in the cosmological context.
In particular, we simulate a suite of isolated galaxy models that have
the same Toomre $Q$ parameter and swing amplification parameter but
that differ in the vertical scale height and velocity dispersion.  We
find that the onset of bar formation occurs more slowly in models with
thicker discs.  Moreover, thicker discs and discs in
simulations with larger force softening are found to be more
resilient to buckling, which acts to regulate the length and strength
of bars.  We  simulate disc-halo systems in the cosmological
environment using a disc-insertion technique developed in a previous
paper.  In this case, bar formation is driven by the stochastic
effects of a triaxial halo and subhalo-disc interactions and the
initial growth of bars appears to be relatively insensitive to the
thickness of the disc.  On the other hand, thin discs in cosmological
haloes do appear to be more susceptible to buckling than thick ones and
therefore bar strength correlates with disc thickness as in the
isolated case.  Thus, one can form discs in cosmological
simulations with relatively weak bars or no bars at all provided the
discs as thin as the discs we observe and the softening length is
smaller than the disc scale height.

@@

The problem of bar formation in disc galaxies tests our understanding
of cosmological structure formation and galactic dynamics.  In
principle, theories of galaxy formation should yield predictions for
the fractional distribution of bars in terms of their strength,
length, and pattern speed.  While it is often difficult to make
precise, quantitative statements about bars from observations, general
properties of their distribution have emerged (See
\citet{sellwood1993}, \citet{Sellwood2013} and \citet{BT} for
reviews).  Roughly 25-30 per cent of disc galaxies exhibit strong
bars {(that is bars that dominate the disc luminosity) and another 20 per
cent or more have relatively weak bars \citep{bar_fraction_2,bar_fraction_1, masters2010, Sellwood2013}. } The bar fraction appears to
increase with time.  Approximately one tenth of disc galaxies between
$0.5 \le z \le 2$ have visually identifiable strong bars, which is a
factor of 3-4 smaller than the fraction in the local Universe (see
\citet{simmons2014} and references therein).

The bar fraction also
varies with galaxy type.  \citet{masters2010} find that $70\pm 5$ per
cent of the so-called passive red spirals have bars as compared to a
$25\pm 5$ per cent bar fraction for blue spirals.  Since the red
spirals are interpreted as old galaxies that have used up their
star-forming gas, this result is consistent with a bar fraction that
increases with time.

{Further evidence of a colour correlation with galaxy bar fractions was presented in \citet{masters2011}. They also found an increased bar
fraction in galaxies with dominant bulges. Other correlations were explored by \citet{na2010}, who found correlations between bar fractions galaxies of different morphological types and characteristic masses. They also found that central mass concentrations were correlated with different strong bar fractions depending on the morphological type and mass of the galaxies considered. In the same line of thought, \citet{masters2012} found a correlation between the gas content of a galaxy and the likelihood of a strong bar being present.}

{Putting all of these observations together}, 
disc galaxies in the local Universe divide into three
roughly equal bins: those with strong bars, those with weak bars, and
those with no detectable bar.  {These observations suggest that
bars are capable of forming at a wide range of cosmic times, but 
once formed are difficult to destroy. Though bars can, in theory, be
destroyed through an increased central mass concentration, the requisite
mass exceeds typical masses for supermassive black holes \citep{athanassoula_2005}.
}

%\textbf{These observations suggest that bars are 
%capable of forming at a wide range of cosmic times, but once formed,
%are difficult to destroy. Bars can, in theory, be destroyed or weakened 
%by growing a large central mass, but this mass often exceeds typical masses for
%central supermassive black holes \citep{athanassoula_2005}.}

Intuition tells us that properties of a bar should depend on the
properties of its host galaxy and the environment in which that
galaxy lives.  Theoretical arguments indicate that cold, thin discs
are susceptible to local ``Toomre'' instabilities {\citep{BT}}.  
Furthermore, discs that are strongly self-gravitating, that is discs that 
contribute the bulk of the gravitational force required to maintain their rotational
motion, are susceptible to global instabilities.  Thus, one can
construct initially axisymmetric galaxy models that form bars with
vastly different properties (or no bars at all) by changing the
internal disc dynamics or trading off disc mass for bulge and halo
mass.  The implication is that the {structure of bars} provides an
indirect means for inferring a disc's kinematics and mass-to-light
ratio as well as the distribution of mass in a galaxy's dynamically
``hot'' components, namely its bulge and dark matter halo. {However,
it is difficult to infer the mass distribution of a galaxy from bar
strengths alone since it depends on the halo velocity dispersion, 
which cannot be observed directly \citep{athanassoula_2003}.}

%\textbf{However,
% as \citet{athanassoula_2003} notes, we run into ambiguity while
%inferring mass distributions from bar strengths alone, since these depend
%on an unobservable halo velocity dispersion.}

A galaxy's ability to resist local instabilities is typically
expressed in terms of the (kinetic) Toomre $Q$-parameter {(see Eq. \eqref{eq:q})}
\citep{ToomreParameter} while its ability to resist global
instabilities is encapsulated in the swing-amplification $X$-parameter {(see Eq. \eqref{eq:xm})}
\citep{GoldreichTremaine1978,GoldreichTremaine1979}.  Both parameters
are defined so that large values imply a more stable disc.  The
hypothesis that they determine a galaxy's susceptibility to bar
formation has been tested by simulations of isolated, idealized galaxy
models \citep{PeeblesOstriker1973, ZangHohlBars1978,
  CombesSandersBars1981,Sellwood1981}.  Typically, the initial galaxy
is represented by an N-body {(Monte Carlo)} realization of an equilibrium solution to
the collisionless Boltzmann equation comprising a disc, dark matter
halo, and often, a bulge.  Equilibrium does not imply
stability, and a galaxy can develop spiral structure and a bar through
instabilities that are seeded by shot noise
\citep{EfstathiouShotNoise} and amplified by feedback loops such as
swing amplification \citep{Sellwood2013}. {A common way to suppress the mechanisms
which give rise to these effects is} by increasing either $Q$ or $X$.  For
example, in dynamically warm discs (that is, discs with high $Q$)
perturbations are randomized on timescales short enough to prevent the
feedback loop from starting \citep{AthanassoulaSellwood1986}.
Likewise, submaximal discs, that is, discs with high $X$, avoid the
bar instability presumably because the disc lacks the self-gravity to
drive the bar instability \citep{EfstathiouShotNoise,
  ChristodoulouStability1995, Sellwood2013}.

@@

Of course, galaxies are neither isolated nor born as axisymmetric,
equilibrium systems. 
%\textbf{In these idealized systems, instabilities can only come from shot noise, whereas this is is not true in a complicated dynamical environment. As such,} 
{In N-body realizations of these idealized systems, instabilities
are seeded from the shot noise of the particle distribution. On the other
hand, interactions in real galaxies can be driven by interactions
with the given galaxy's environment. As such, the process of} bar formation may be very different in idealized
galaxies as compared with galaxies in a cosmological setting.  For
example, halo substructure in the form of satellite galaxies and dark
matter subhaloes can pass through and perturb the disc.
\citet{gauthier2006}, \citet{kazantzidis2008}, and
\citet{dbs2009} showed that an apparently stable disc galaxy
model can develop a bar when a fraction of the ``smooth'' halo is
replaced by substructure in the form of subhaloes.  The effect is
stochastic; subhalo-triggered bar formation seems to require subhaloes
whose orbits take them into the central regions of the disc in a
prograde sense.  More recently, \citet{purcell2011} showed that
Sagittarius dwarf alone could have been responsible for the Milky
Way's spiral structure and bar.

@@

Though the \citet{YurinSpringelStellarDisks} models vary in $Q$ and
$X$ they share two important properties.  First, the ratio of the
radial and vertical velocity dispersion is set to unity throughout the
disc.  Second, the ratio of the vertical and radial scale lengths is
fixed at $0.2$, which is roughly a factor of two larger than that of
the Milky Way's thin disc.  In addition, the gravitational softening length in their
simulations is fixed at $680\,{\rm pc}$.  Thus, the discs in their
simulations are thicker and (vertically) warmer than what one might
expect for Milky Way-like galaxies.  The results from
\citet{Klypin2009} suggest that these properties rather than (or
together with) halo substructure and triaxiality are responsible for
the preponderance of strong bars in the
\citet{YurinSpringelStellarDisks} models.

@@

We begin in section \S \ref{sec:ics} by discussing the dimensionless
parameters that arise when one constructs equilibrium disc galaxy
models.  These include the aforementioned $Q$ and $X$ parameters as
well as the ratio of the vertical and radial velocity dispersion and
the ratio of the vertical and radial scale lengths.  We also discuss
possible effects of gravitational softening.  In
\ref{sec:thick_discs_suppress} we describe our sequence of simulations
and discuss how they fit in with previous work.  We discuss results
for our isolated galaxy simulations in \S \ref{sssec:isolated} and our
cosmological ones in \S \ref{sec:cosmo}.  We conclude with a
discussion of the implications of our results in \S
\ref{sec:conclusions}.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{../figures/rotation_curves_two_panel.eps}
	\caption{Rotation curve decomposition for our models. Total
          rotation curves are shown as solid lines while the separate
          contributions from the disc and halo are shown as dashed and
          dot-dashed curves, respectively.  Blue curves in the top
          panel are for the isolated galaxy simulations with
          \textsc{GalactICS} initial conditions while the green curves
          are for the simulations C.I.Ag run with \textsc{AGAMA}
          initial conditions.  Bottom panel shows initial rotation
          curve decomposition for the runs D.I and
          E.II.}\label{fig:rcs}
\end{figure}

\begin{sidewaystable}
\begin{tabular}{l l l l l l l l l l l l l}
\hline
 & $M_d$ & $R_d$ & $V_c$ & $\sigma_R$ & $z_d/R_d$  & $\sigma_R/\sigma_z$  & $X$ &  $Q$ & $\rho_h/\rho_0$ & $\epsilon$ & {Description}\\ 
\hline
A.I{/A.III}     & 3.49 & 2.50 & 216 & 25.3 & 0.10 & 1.27 & 2.34 & 1.00 & 0.14 & 0.15 & {Thin, Fiducial / Thin, Isotropized}\\
A.II    & 3.49 & 2.50 & 216 & 25.3 & 0.10 & 1.27 & 2.34 & 1.00 & 0.14 & 0.50 & {Thin, High Softening}\\
B.I{/B.III}     & 3.49 & 2.50 & 213 & 25.3 & 0.20 & 0.97 & 2.34 & 1.00 & 0.28 & 0.15 & {Mid, Fiducial / Mid, Isotropized}\\
B.II    & 3.49 & 2.50 & 213 & 25.3 & 0.20 & 0.97 & 2.34 & 1.00 & 0.28 & 0.50 & {Mid, High Softening}\\
C.I     & 3.49 & 2.50 & 208 & 25.3 & 0.40 & 0.77 & 2.34 & 1.00 & 0.50 & 0.15 & {Thick, Fiducial} \\
C.I.Ag  & 3.49 & 2.50 & 216 & 25.3 & 0.40 & 0.77 & 2.34 & 1.00 & 0.50 & 0.15 & {Thick, AGAMA ICs}\\
D.I    & 5.82 & 3.70 & 245 & 25.2 & 0.10 & 1.14 & 2.45 & 1.00 & 0.29 & 0.18  & {Thin, Cosmological}\\
E.II    & 5.82 & 3.70 & 245 & 27.4 & 0.25 & 0.72 & 2.45 & 1.00 & 0.73 & 0.74 & {Thick, High Softening, Cosmological}\\
\hline
YS15.A5 & 5.00 & 3.00 & 263 & 30.7  & 0.2 & 1.00 & 3.22 & 1.38 & 0.21 & 0.68 & {--}\\
YS15.B5 & 5.00 & 3.00 & 211 & 26.6  & 0.2 & 1.00 & 2.06 & 0.96 & 0.11 & 0.68 & {--}\\
YS15.C5 & 5.00 & 3.00 & 270 & 30.3  & 0.2 & 1.00 & 3.31 & 1.42 & 0.23 & 0.68 & {--}\\
YS15.D5 & 5.00 & 3.00 & 236 & 26.6  & 0.2 & 1.00 & 2.58 & 1.12 & 0.16 & 0.68 & {--}\\
YS15.E5 & 5.00 & 3.00 & 233 & 27.1  & 0.2 & 1.00 & 2.58 & 1.11 & 0.15 & 0.68 & {--}\\
YS15.F5 & 5.00 & 3.00 & 219 & 27.0  & 0.2 & 1.00 & 2.22 & 1.02 & 0.11 & 0.68 & {--}\\
YS15.G5 & 5.00 & 3.00 & 227 & 28.2  & 0.2 & 1.00 & 2.45 & 1.09 & 0.13 & 0.68 & {--}\\
YS15.H5 & 5.00 & 3.00 & 244 & 28.6  & 0.2 & 1.00 & 2.85 & 1.21 & 0.16 & 0.68 & {--}\\
\hline
G06     & 7.77 & 5.57 & 226 & 17.1  & 0.06  & 1.80 & 2.78 & 1.43 & 0.10 & 0.15 & {--}\\
\hline
%$M_d$ & $7.2 \times 10^{10} M_\odot$\\
%$N_d$ & $10^6$\\
%$R_{d,0}$ & 3.7 kpc\\
%$N_{r}$ & $4096^3$
\end{tabular}
\caption{Summary of parameters for the simulations considered in this
  paper, the disc-halo simulations considered in
  \citet{YurinSpringelStellarDisks} (labeled YS15) and the
  \citet{gauthier2006} (G06).  $M_d$ is the final disc mass in units of
  $10^{10}\,M_\odot$ , $R_d$ is the disc scale radius in units of
  ${\rm kpc}$, and $V_c$ and $\sigma_R$ are the circular speed and
  radial velocity dispersion in units of ${\rm km\,s^{-1}}$ and
  evaluated at $R_p = 2.2R_d$.  For the disc aspect ratio, we quote
  $z_d/R_d$ where $z_d$ is the ${\rm sech}^2$-scale length.  The
  velocity dispersion ratio $\sigma_R/\sigma_z$, the $X$ and $Q$
  parameters, the ratio of the halo density in the midplane to that of
  the disc, and the logarithmic derivative of the circular speed are
  also measured at $R_p$.  Finally, the softening length $\epsilon$ is
  given in units of ${\rm kpc}$.  Simulations A.III and B.III are the
  same as A.I and B.I except that they are run with vertical motions
  isotropized so as to shut off the buckling
  instability.} \label{tab:sims}
\end{sidewaystable}

@@

\begin{figure}
		\centering
        \includegraphics[width=0.8\textwidth]{../figures/falpha.eps}
        \caption{The dimensionless ratio $\sigma_z^2/\pi G\Sigma z_d$ as a
          function of $\rho_h/\rho_0$ for the models considered in
          this paper (stars), the disc-halo models from
          \citet{YurinSpringelStellarDisks} (filled squares) and the
          model from \citet{gauthier2006} (filled triangle).  The
          straight line is the function $f = 1 + \left (2\pi/3\right
          )^{1/2}\rho_h/\rho_0$ discussed in the text.}
\label{fig:falpha}\end{figure}

\begin{figure}
		\centering
        \includegraphics[width=1.\textwidth]{../figures/qxr.eps}
        \caption{Distribution of simulations considered in this paper
          in the $Q-X$ and the $z_d/R_d-\sigma_R/\sigma_z$ planes.
          Stars are simulations run for this paper (A-E); filled
          squares denote the series of simulations described in
          \citet{YurinSpringelStellarDisks}; the filled triangle denotes
          the simulation of M31 run in \citet{gauthier2006}; open
          pentagons denote the simulations described in
          \citet{WPDGalactICSReference}.
        }\label{fig:QXR}\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{../figures/wpd.eps}
	\caption{Strength and length of bars for the simulations
          considered in \citet{WPDGalactICSReference}.  The twenty-five models
          span the $Q$-$X$ plane.  Top panel shows the $A_2$ parameter
          while the bottom panel shows the bar length, $l_b$.  Both are
          measured at $5\,{\rm Gyr}$ (the final snapshot of the
          simulations). {The bar length is the full bar length.}}
	\label{fig:qxa2}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../figures/isolated_xy_with_agama_with_circles.eps}
	\caption{Surface density maps for isolated galaxy simulations
          at select times. Time proceeds from 0 to 10 Gyr,
          left-to-right, and the models span top-to-bottom in order of
          their appearance in Table \ref{tab:sims}. The overlaid red
          circles have radii $R_p=5.5\,{\rm kpc}$ and 20 kpc. {Recall $R_P=2.2 R_d$, the 
          radius of peak circular velocity.}}
	\label{fig:face_on_isolated}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{../figures/isolated_a2_vs_t_2rd_weighted.eps}
	\caption{Mean bar strength parameter inside a cylindrical
          radius $R_p$, $A_2(<R_p)$, as a function of time. Curves are
          smoothed in time with a top-hat moving window of width
          $1\,{\rm Gyr}$.  Line colours are blue, red, and {orange} for
          models A, B, and C, respectively.  Results for the
          fiducial runs A.I, B.I, and C.I are shown as solid curves
          while the results for the runs with high softening length,
          A.II and B.II, are shown as dashed curves.  The
          \textsc{AGAMA} model C.I.Ag is shown as a dot-dashed
          curve.} \label{fig:isolated_a2_vs_t}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.1\textwidth]{../figures/isolated_r_t_a2_with_agama.eps}
	\caption{Bar strength parameter $A_2$ as a function of radius
          and time.  The trajectory of corotation is shown by the
          dashed red line.} \label{fig:isolated_r_t_a2}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../figures/isolated_four_panel_zrms.eps}
	\caption{Root mean square height $\zrms$ as a function of
          cylindrical radius $R$ for ten snapshots equally spaced over
          the first $500\,{\rm Myr}$.  Panels are for simulations A.I
          (upper left), A.II (upper right), C.I (lower left) and
          C.I.Ag (lower right).} \label{fig:zrms}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../figures/isolated_dispersion_evolution_with_agama_zrms.eps}
	\caption{Diagonal components of the velocity dispersion tensor
          and $\zrms$ as a function of $R$ for different snapshots
          between $0$ and $10\,{\rm Gyr}$.  Shown, from to to bottom,
          are profiles for $\zrms$, $\sigma_z$, $\sigma_R$, and $\sigma_\phi$
          for the same size models included in
          Fig.\,\ref{fig:face_on_isolated}.} \label{fig:isolated_dispersions}
\end{figure}

@@

\subsection{Initial Conditions for Isolated Galaxy Simulations} \label{sssec:isolated}
